\chapter{Results and Discussion}
\label{chap:eval}

This chapter discusses the results of the study.
Section \ref{chap:eval:sec:results} summarizes the results and contribution of this study, and outlines the direction for further studies. Section \ref{chap:eval:sec:limitations} discusses the limitations of the study. Section \ref{chap:eval:sec:further_work} outlines the further work in the direction of this research. Section \ref{chap:eval:sec:conclusion} concludes the chapter.

\section{Simulation Results}
\label{chap:eval:sec:results}
RRMGreedy++ is shown to both RRMGreedy and LCCS in terms of convergence rate, SNR, and throughput.

\section{Limitations}
\label{chap:eval:sec:limitations}

Using the simulated environment provides a controlled and reproducible environment for evaluation, however, networking and physical signal propagation models used pose their constraints. For example, \texttt{YansWifiPhy} propagation model does not consider cross-channel interference between adjacent channel, which can significantly impact channel planning, especially on 2.4 GHz band.

\section{Further Work}
\label{chap:eval:sec:further_work}
As comes from the limitations (Section \ref{chap:eval:sec:limitations}), I can identify two directions for the following research:
\begin{enumerate}
    \item Improving the simulation model: the next steps would be to extend the simulation environment to consider more realistic scenarios. This includes employing cross-channel interference model SpectrumWifiPhy, considering the impact of hidden nodes, and introducing a wireless LAN controller (WLC) as a part of the simulated network, since current implementation works "above" the nodes, being a part of the script logic;
    \item Experimenting with another channel selection approaches: namely, I consider channel planning using genetic algorithms \cite{raschellaEvaluationChannelAssignment2019} to be a prospective solution with performance on-par with exact optimization-based RRM algorithms \cite{raschellaEvaluationChannelAssignment2019};
    \item Implementing the RRMGreedy algorithm in a real-world test bed: since the algorithm aligns with constraints described in Section \ref{chap:lr:sec:conclusion}, metrics required for its operation can be obtained from the majority of OpenWRT-based Wi-Fi access points available on the market.
\end{enumerate}

\section{Conclusion}
\label{chap:eval:sec:conclusion}
In this thesis, I derived an improved version of \textit{RRMGreedy} Radio Resource Management (RRM) algorithm, \textit{RRMGreedy}, and implemented evaluation of these algorithms using ns-3 network simulator.
To achieve this, I performed analysis of \textit{RRMGreedy}, and identified its flaws.
% I have also implemented a local per-cell RRM algorithm, Least Congested Channel Search (LCCS), as a baseline for comparison.
Moreover, since ns-3 does not provide facilities for implementing RRM algorithms out of the box, and since Wi-Fi station channel switching logic was not implemented as of current ns-3 release, I have introduced a novel RRM framework for ns-3 and a set of enhancements into the existing codebase.
Finally, the algorithms were evaluated using various simulation settings. The results of the simulation show that \textit{RRMGreedy++} outperforms \textit{RRMGreedy} in terms of SNR, throughput, and spectrum management. The further work in this direction would be making the simulation model more accurately represent real-world deployments, experimenting with optimization approaches and heuristics other than greedy search, and evaluating \textit{RRMGreedy} on a real-world test bed.